---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `netseg`: Measures of Network Segregation and Homophily

<!-- badges: start -->
<!-- badges: end -->

This package implements most important descriptive measures of homophily and segregation in social networks. See Bojanowski & Corten (2014) for a review.

## Installation

<!--
You can install the released version of netseg from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("netseg")
```
-->

The development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("mbojan/netseg")
```







## Example

To illustrate, consider the network below which comes from White (1975)


```{r data-white}
library(netseg)
library(igraph)

data("WhiteKinship")

plot(
  WhiteKinship,
  vertex.color = "white",
  vertex.shape = c("circle", "square")[match(V(WhiteKinship)$gender, c("female", "male"))]
)
```

The extent of segregation in this network can be assessed using one of the indices provided in the package, for example:

```{r segregation}
# network-level
freeman(WhiteKinship, "gender")

# node-level
ssi(WhiteKinship, "gender")
```




## References

Bojanowski, M., & Corten, R. (2014). Measuring segregation in social networks. *Social Networks*, 39, 14-32.

White, D. R. (1975) "Communicative Avoidance in Social Networks". University of California, Irvine. (mimeo) 
